\section*{Overview}

Due to technological limitations, it is proving increasingly difficult to
maintain a continual increase in the performance of individual processors.
Therefore, the current trend is to integrate multiple processors on to a single
chip and exploit the resulting parallel resources to achieve higher computing
power. However, this may require significantly different approaches to both
hardware and software particularly for general purpose applications. This course
will explore these issues in detail.

\section*{Syllabus}
\begin{mymulticols}
\begin{description}
\item{\textbf{Introduction}} Trends in technology, limitations and consequences. The move to multi-core parallelism in programs, ILP, Thread Level, Data Parallelism.
\item{\textbf{Parallel Architectures}} SIMD, MIMD, Shared Memory, Distributed Memory, strengths and weaknesses.
\item{\textbf{Parallel Programming}} Multithreaded programming, Data parallel programming, Explicit vs Implicit parallelism, automatic parallelisation. The Case for Shared Memory. When to share?
\item{\textbf{Shared Memory Multiprocessors}} Basic structures, the cache coherence problem. The MESI protocol. Limitations. Directory based coherence.
\item{\textbf{Programming with Locks and Barriers}} The need for synchronisation. Problems with explicit synchronisation
\item{\textbf{Other Parallel Programming Approaches}} MPI and OpenMP
\item{\textbf{Speculation}} The easy route to automatic parallelisation?
\item{\textbf{Transactional Memory}} Principles. Hardware and Software approaches
\item{\textbf{Memory Issues}} Memory system design. Memory consistency
\item{\textbf{Other Architectures and Programming Approaches}}
GPGPUs, CUDA
\item{\textbf{Data Driven Parallelism}} Dataflow principles and Functional Programming
\end{description}
\end{mymulticols}

\section*{Attribution}

These notes are based off of both the course notes
(\url{http://studentnet.cs.manchester.ac.uk/ugt/2015/COMP35112/}). Thanks to John Gurd for such a good course! If you find any errors, then I'd love to hear about them!

\section*{Contribution}

Pull requests are very welcome:
\url{https://github.com/Todd-Davies/third-year-notes}